Eco-evolutionary dynamics of finite populations from demographic first principles

### 2023-01-21 ind_files/03_BD_processes.tex
Add example for multidimensional processes

@@

@@

from which it is clear that we have obtained a continuous time analog of the Price equation.

@@

        $1 < m < \infty$ \newline (Discrete traits) & $[0,1,2,3,\ldots]^{m}$ \newline (Number of individuals of each trait variant) & $2m$ real-valued functions, $b_i(\mathbf{v})$ and $d_i(\mathbf{v})$ (for $1 \leq i \leq m$) describing the birth and death rate of trait variant $i$ when the population is $\mathbf{v}$ & Multivariate Fokker-Planck equation \newline ($m$-dimensional SDE) &  Evolutionary game theory \newline Quasispecies equation \newline Price equation (discrete traits) \\

>>>>

\label{1D_det_limit}

@@

\label{nD_WNA_directional_derivative_for_replicator_eqns}

@@

from which it is clear that we have obtained a version of the Price equation for quantitative traits.

@@

        $1 < m < \infty$ \newline (Discrete traits) & $[0,1,2,3,\ldots]^{m}$ \newline (Number of individuals of each trait variant) & $2m$ real-valued functions, $b_i(\mathbf{v})$ and $d_i(\mathbf{v})$ (for $1 \leq i \leq m$) describing the birth and death rate of trait variant $i$ when the population is $\mathbf{v}$ & Multivariate Fokker-Planck equation \newline ($m$-dimensional SDE) &  Evolutionary game theory \newline
        Lotka-Volterra competition \newline Quasispecies equation \newline Price equation (discrete traits) \\

### 2023-01-21 ind_files/04_polymorphisms.tex
Add example for multidimensional processes

\section{Example in one dimension: The stochastic logistic equation}
Consider the functional forms of example \ref{ex_1D_stoch_logistic}, given by
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
    \frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. The noise is controlled by
\begin{equation*}
    \sqrt{\frac{A^+(x)}{K}} = \sqrt{\frac{x(v+rx)}{K}}
\end{equation*}
and thus, the diffusion approximation (`mesoscopic view') of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dB_t
\end{equation}
Letting $\alpha(t)$ be the solution of the logistic equation \eqref{ex_1D_stoch_logistic_det_limit}, We can Taylor expand $A^{\pm}(x)$ for the weak noise approximation, and we find:
\begin{align*}
    A^-_1(x) &= \frac{d}{dx}(rx(1-x))\biggl{|}_{x=\alpha} = r(1 - 2\alpha(t))\\
    A^+_0(x) &= \alpha(t)(v+r\alpha(t))
\end{align*}
Thus, the weak noise approximation of \ref{ex_1D_stoch_logistic_BD_eqns} is given by
\begin{equation}
    X_t = \alpha(t) + \frac{1}{\sqrt{K}}Y_t
\end{equation}
where the stochastic process $Y_t$ is an Ornstein-Uhlenbeck process given by the solution to the linear SDE
\begin{align}
    dY_t &= A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dB_t\nonumber\\
    \Rightarrow dY_t &= r(1 - 2\alpha(t))Y_tdt + \sqrt{\alpha(t)(v+r\alpha(t))}dB_t\label{ex_1D_stoch_logistic_WNA}
\end{align}

@@

>>>>

\section{Example in one dimension: The stochastic logistic equation}
Consider the functional forms of example \ref{ex_1D_stoch_logistic}, given by
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dB_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we
obtain the logistic equation. Letting $\alpha(t)$ be the solution of the logistic equation \eqref{ex_1D_stoch_logistic_det_limit}, We can Taylor expand $A^{\pm}(x)$ for the weak noise approximation, and we find:
\begin{align*}
A^-_1(x) &= \frac{d}{dx}(rx(1-x))\biggl{|}_{x=\alpha} = r(1 - 2\alpha(t))\\
A^+_0(x) &= \alpha(t)(v+r\alpha(t))
\end{align*}
Thus, the weak noise approximation of \ref{ex_1D_stoch_logistic_BD_eqns} is given by
\begin{equation}
    X_t = \alpha(t) + \frac{1}{\sqrt{K}}Y_t
\end{equation}
where the stochastic process $Y_t$ is an Ornstein-Uhlenbeck process given by the solution to the linear SDE
\begin{align}
    dY_t &= A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dB_t\nonumber\\
    \Rightarrow dY_t &= r(1 - 2\alpha(t))Y_tdt + \sqrt{\alpha(t)(v+r\alpha(t))}dB_t\label{ex_1D_stoch_logistic_WNA}
\end{align}

@@

\section{Example for discrete traits: Lotka-Volterra and matrix games in finite populations}

The methods outlined in the above section have recently been used to study the population dynamics of a finite population playing a so-called `matrix game' (An evolutionary game for which you can write down a payoff matrix) with 2 pure strategies \citep{tao_stochastic_2007}. Based on the interpretation of what each type represents, this is mathematically equivalent to studying frequency-dependent selection on a one-locus two-allele gene (with a bijective genotype-phenotype map and no mutations) or studying two-species competitive Lotka-Volterra dynamics, as we will show below.\\
\\
Let us imagine a population with $m$ types of individuals that are competing for resources. Let the state of the population be characterized by the vector $\mathbf{v}(t) = [v_1(t),v_2(t),\ldots,v_m(t)]$, where $v_i(t)$ is the number of type $i$ individuals at time $t$. Let the birth and death rates of the $i$th type be given by:
\label{nD_example_numbers_b_d_rates}
b_i(\mathbf{v}) &= \lambda v_i\\
d_i(\mathbf{v}) &= \left(\mu + \frac{1}{K}\left(\sum\limits_{j=1}^{m}M_{ij}v_j\right)\right)v_i
where $K > 0$ is our system size parameter (and represents a global carrying capacity across all types), $\lambda > 0$ and $\mu > 0$ are suitable positive constants representing the baseline natality and mortality common to all types, and $M_{ij} \geq 0$ is a non-negative constant describing the effect of type $j$ individuals on the death rate of type $i$ individuals due to competition. We assume that $M_{ij} \ll K$. The values $M_{ij}$ are often collected in an $m \times m$ matrix $\mathbf{M}$. The negative of this matrix, $-\mathbf{M}$, is called the `payoff matrix' (in evolutionary game theory) or `interaction matrix' (in Lotka-Volterra models), which is why the infinite population analogues of such models are called `matrix games' in the evolutionary game theory literature. Lotka-Volterra models also frequently assume that the diagonal elements $M_{ii}$ are all equal, though we will not make that assumption here.\\
\\
Going from population numbers $\mathbf{v}$ to densities $\mathbf{x} = \mathbf{v}/K$, we obtain the birth and death rates:
\label{nD_example_density_b_d_rates}
b^{(K)}_i(\mathbf{x}) &= \lambda x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}M_{ij}x_j\right)x_i
Thus, we have
A^{\pm}_{i} = x_i\left(\lambda \pm \left(\mu + \sum\limits_{j=1}^{m}M_{ij}x_j\right)\right)
Defining $r = \lambda - \mu$ and $v = \lambda + \mu$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r - \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
and
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(v + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
From \eqref{nD_det_limit}, we see that the deterministic limit is a set of $m$ coupled ODEs given by
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_i\left(r - \sum\limits_{j=1}^{m}M_{ij}x_j\right)
These are precisely the competitive Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\label{nD_example_fitness}
b^{(\textrm{int})}_{i}(\mathbf{x}) &= \lambda\\
d^{(\textrm{int})}_i(\mathbf{x}) &= \mu + \sum\limits_{j=1}^{m}M_{ij}x_j\\
w_i(\mathbf{x}) &= r - \sum\limits_{j=1}^{m}M_{ij}x_j
If $p_i(t)$ is the frequency of type $i$ individuals in the population at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r - \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r - \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\frac{dp_i}{dt} = \left[\sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right) - \sum\limits_{j=1}^{m}M_{ij}x_j\right]p_i
Let us define the payoff matrix $\mathbf{B} = - \mathbf{M}$. Then, the dynamics are
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Bp})_i - \mathbf{p}\cdot\mathbf{Bp}\right]p_i   
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $B$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks.\\
Let the solution to the equations \eqref{nD_example_det_limit} be given by $\mathbf{a}(t) = [a_1(t),\ldots,a_m(t)]$. For the weak noise approximation, we can Taylor expand $A^{\pm}_i$ and use \eqref{nD_WNA_directional_derivative_for_replicator_eqns} to compute the directional derivative as:
D_i &= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial w_i}{\partial x_k}\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_k\left(\frac{\partial}{\partial x_k}(r+\sum\limits_{j=1}^{m}B_{ij}x_j)\bigg{|}_{\mathbf{x}=\mathbf{a}(t)}\right)\\
&= y_iw_i(\mathbf{a}) + a_i\sum\limits_{k=1}^{m}y_kB_{ik}\\
\Rightarrow D_i &= y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\label{nD_example_directional_derivative}
where we have used the fact that $w_i(\mathbf{y}) = r + \sum\limits_{k=1}^{m}y_kB_{ik}$ (from \eqref{nD_example_fitness}) in the last step. Thus, in the weak noise approximation of our process, the dynamics are given by
\mathbf{x}(t) = \mathbf{a}(t) + \frac{1}{\sqrt{K}}\mathbf{y}(t)
where the stochastic fluctuations $\mathbf{y}(t)$ satisfy the linear Fokker-Planck equation
\resizebox{1.1\textwidth}{!}{$\displaystyle\frac{\partial P}{\partial t}(\mathbf{y},t) = \sum\limits_{i=1}^{m}\left(-\frac{\partial}{\partial y_i}\left\{\left(y_iw_i(\mathbf{a}) + a_iw_i(\mathbf{y}) - ra_i\right)P(\mathbf{y},t)\right\}+\frac{1}{2}\left(a_i\left(v - \sum\limits_{j=1}^{m}B_{ij}a_j\right)\right)\frac{\partial^2}{\partial{y_i}^2}P(\mathbf{y},t)\right)$}
Using \eqref{nD_example_directional_derivative} in \eqref{nD_moment_eqn_mean}, we see that the fluctuations are expected to evolve as:
\label{nD_example_moment_eqn_mean}
\frac{d}{dt}\mathbb{E}[y_i] = w_i(\mathbf{a})\mathbb{E}[y_i] + a_i\sum\limits_{k=1}^{m}B_{ik}\mathbb{E}[y_k]
% or, in matrix form:
% \begin{equation}
% \resizebox{0.91\textwidth}{!}{$\displaystyle
% \frac{d}{dt}\begin{bmatrix}
%     \mathbb{E}[y_1]\\
%     \mathbb{E}[y_2]\\
%     \vdots\\
%     \mathbb{E}[y_i]\\
%     \vdots\\
%     \mathbb{E}[y_m]
%     \end{bmatrix}
%     =
% \begin{bmatrix}
%     (r + \sum\limits_{j=1}^{m}B_{1j}a_j + a_1B_{11}) & a_1B_{12} & a_1B_{13} & \dots & \dots & \dots & a_1B_{1m}\\
%     a_2B_{21}) & (r + \sum\limits_{j=1}^{m}B_{2j}a_j + a_2B_{22}) & a_2B_{23} & \dots & \dots & \dots & a_2B_{2m}\\
%     \vdots & \vdots & \vdots &\vdots & \vdots & \vdots & \vdots\\
%     a_{i}B_{i1} & a_iB_{i2} & a_iB_{i3} & \dots & (r + \sum\limits_{j=1}^{m}B_{ij}a_j + a_iB_{ii}) & \dots & a_iB_{im}\\
%     \vdots & \vdots & \vdots &\vdots & \vdots & \vdots & \vdots\\
%     a_mB_{m1} & a_mB_{m2} & a_mB_{m3} & \dots & \dots & \dots & (r + \sum\limits_{j=1}^{m}B_{mj}a_j + a_mB_{mm})
%     \end{bmatrix}
% \begin{bmatrix}
%     \mathbb{E}[y_1]\\
%     \mathbb{E}[y_2]\\
%     \vdots\\
%     \mathbb{E}[y_i]\\
%     \vdots\\
%     \mathbb{E}[y_m]
%     \end{bmatrix}
% $}
% \end{equation}
\hl{can compare time series or something here if we think it is helpful. Visualization may be little hard though, because we always have $\geq 3$ dimensions ($m$, + 1 additional for time)}
